---
title: "test report on lepidoptera"
author: "Dalon White"
date: "2025-10-15"
output: html_document

params:
  begin_date: NULL #NULL for all dates, otherwise "mm-dd-yyyy"
  end_date: "today" #NULL for today, otherwise "mm-dd-yyyy"
  
  # Commodity taxonomy:
  commodity_taxonomic_name: ["Begonia"] #NULL or ["taxonomic name1", "taxonomic name2", "etc."]
  commodity_common_name: ["Begonia"]
  
  # Country of origin:
  origin_country: NULL #NULL or ["country1", "country2", "etc."]
  
  # Pest taxonomy:
  KINGDOM: NULL #["Fungi"]
  PHYLUM: NULL #["Phylum2", "Phylum3", "etc."]
  CLASS: NULL #["Insecta", "Class2", "etc."]
  ORDER: NULL #["Coleoptera", "Order2", "etc."]
  FAMILY: NULL #["Dermestidae", "Family2", "etc."]
  GENUS: NULL #["Genus2", "Genus3", "etc."]
  SPECIES: NULL #["Granarium", "Species2", "etc."]

  # For the title of the file on save:
  pest_nickname: ["any pest"]
  
---
# If errors:
If VS Code cannot knit, it cannot find your params; run in RStudio instead

# Load libraries and parameters
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load packages
pacman::p_load(DBI, odbc)
pacman::p_load(tidyverse, lubridate, readr, stringr,stringdist)
pacman::p_load(DT, htmlwidgets)

```


```{r load libraries and grab params}
source(here::here("functions","filter_dates.R"))
begin_date <- get_date_params(params=params)[1]
end_date <- get_date_params(params=params)[2]

```


# Gather data from ARM

## Connect to ARM
```{r connect to ARM}

source(here::here("functions","connect_and_access_arm.R"))
#connect to ARM
db_conn <- connect_db()

```

# retrieve diagnostic results table

```{r grab arm results}
source(here::here("functions","connect_and_access_arm.R"))
df_diagnostic_results <- mvw_diagnostic_results()
```

# Filter by date
```{r}
# source(here::here("functions","filter_dates.R"))
# # Apply date filters
# df_diagnostic_results <- filter_dates(data = df_diagnostic_results,
#                                       begin_date = begin_date,
#                                       end_date = end_date)
```

# Filter by country
```{r}
source(here::here("functions","filter_country.R"))
df_diagnostic_results <- filter_country(data = df_diagnostic_results, params = params)
```

# Identify columns to collect
```{r grab the arm data}
df_diagnostic_results <- df_diagnostic_results |>   
  dplyr::select(
    INSPECTION_DATE,
    COMMODITY_DISPLAY_NAME,
    COMMODITY_TAXONOMIC_DISPLAY_NAME,
    INSPECTION_PATHWAY,
    ORIGIN,
    SUBCATEGORY,
    INSPECTION_LOCATION_STATE_CODE,
    DETERMINATION_TYPE,
    any_of(starts_with("PEST_TAXONOMIC_")),
    QUARANTINE_RECOMMENDATION,
    DIAGNOSTIC_DETERMINATION_ID,
    DIAGNOSTIC_REQUEST_ID,
    any_of(c(starts_with("number_dead", ignore.case = TRUE),
           starts_with("number_alive", ignore.case = TRUE),
           starts_with("pest_", ignore.case = TRUE))
    )
  )

```

# Collect data

```{r collect data}

arm_results <- df_diagnostic_results |> 
  collect()

```

# Filter by commodity
```{r}
source(here::here("functions","filter_commodity.R"))
arm_results <- filter_commodity(data = arm_results, params = params)

```

# Filter by pest name
```{r}
source(here::here("functions","filter_taxa.R"))
arm_results <- filter_by_taxonomy(arm_results, params, threshold = 0.99)

# Show results
print_matches(arm_results)
```

## Filter interceptions to best known information

```{r correct pest query diagnostic}
source(here::here("functions","parse_diagnostic_determinations.R"))

# get diagnostic records
# combination of diagnostic determinations and diagnostic determination not possible flags
arm_results <- get_diagnostic_results(connection = db_conn, data = arm_results)

df_diagnostic_results <- parse_final_determination(arm_results)

print_diagnostic_results(parsed_determinations_data = df_diagnostic_results)

```

## Write ARM data and Generate Report
```{r write table and report}
setwd(here::here('output','reports'))

# # Create a folder with name based on parameters
# dir_name <- paste0(
#   if(!is.null(params$pest_nickname)) (paste(params$pest_nickname, collapse="_")) else (paste("all pests")),
#   "_on_",
#   if(!(is.null(params$commodity_common_name) | is.null(params$commodity_taxonomic_name)))(paste(
#     if(!is.null(params$commodity_common_name)) (paste(params$commodity_common_name, collapse="_")),
#     if(!is.null(params$commodity_taxonomic_name)) (paste(params$commodity_taxonomic_name, collapse="_")))) else ("anything"),
#   "_from_",
#   if(!is.null(params$origin_country)) (paste(origin_country, collapse="_")) else ("anywhere"),
#   "_from_period_",
#   if(!is.null(params$begin_date)) (params$begin_date) else ("anytime"),
#   "_to_",
#   if(!is.null(params$end_date)) (params$end_date) else ("anytime")
# )
# if(!dir.exists(dir_name)) dir.create(dir_name)

# Generate base filename without extension
base_filename <- paste0(
  if(!is.null(params$pest_nickname)) (paste(params$pest_nickname, collapse="_")) else (paste("all pests")),
  "_on_",
  if(!(is.null(params$commodity_common_name) | is.null(params$commodity_taxonomic_name)))(paste(
    if(!is.null(params$commodity_common_name)) (paste(params$commodity_common_name, collapse="_")),
    if(!is.null(params$commodity_taxonomic_name)) (paste(params$commodity_taxonomic_name, collapse="_")))) else ("anything"),
  "_from_",
  if(!is.null(params$origin_country)) (paste(origin_country, collapse="_")) else ("anywhere"),
  "_from_period_",
  if(!is.null(params$begin_date)) (params$begin_date) else ("anytime"),
  "_to_",
  if(!is.null(params$end_date)) (params$end_date) else ("anytime")
)

# Create interactive data table
dt <- datatable(df_diagnostic_results,
                extensions = 'Buttons',
                options = list(
                  dom = 'Bfrtip',
                  buttons = c('copy', 'csv', 'excel', 'pdf')
                ))

write.csv(df_diagnostic_results, file = (paste0(base_filename, "_data.csv")))


# Save the interactive data table
saveWidget(dt, file = paste0(base_filename, "_data.html"), selfcontained = TRUE)
# Source the generate-report.R file to get the function
source(here::here("add-ons", "generate-report.R"))

# Generate and save the report
report_html <- generate_pest_report(
  data = df_diagnostic_results,
  params = params,
  base_filename = base_filename,
  include_plot = TRUE,
  include_data_preview = TRUE
)

# Save the report to file
report_file <- save_report(report_html, base_filename)

# Print information about saved files
cat(paste0("\nFiles saved to folder: ", getwd(), "/", dir_name, "\n"))
cat(paste0("- Interactive data table: ", base_filename, "_data.html\n"))
cat(paste0("- Search report: ", report_file, "\n"))
```


```{r plot1}
source(here::here('add-ons','plot-diagnostic-results.R'))

diagnostic_results_summary = summarize_diagnostic_results(df_diagnostic_results,
                                       arrange_by = "origin", #or "year"; anything else will be by total pests
                                       truncate_origins = TRUE,
                                       top_n_origins = 10,
                                       group_by_origin = TRUE,
                                       group_by_year = TRUE,
                                       group_by_commodities = FALSE,
                                       truncate_commodities = FALSE,
                                       top_n_commodities = 10,
                                       group_by_taxon = FALSE,
                                       truncate_taxon = FALSE,
                                       top_n_taxon = 10)

save_summary(data = diagnostic_results_summary,
             base_filename = base_filename)




save_plot(data = diagnostic_results_summary,
                      base_filename = paste0(base_filename,'_plot1'),
                      plot_x = "INSPECTION_YEAR",
                      xlab = "Year",
                      plot_y = "total_pests",
                      ylab = "Total Pests",
                      plot_fill = "ORIGIN",
                      fill_lab = "Origin",
                      plot_title = "Pest Interceptions by Year and Origin",
                      plot_subtitle = "Summary of pest interceptions",
                      width = 10,
                      height = 7,
                      dpi = 300)

```

```{r}

diagnostic_results_summary2 = summarize_diagnostic_results(df_diagnostic_results,
                                       arrange_by = "origin", #or "year"; anything else will be by total pests
                                       truncate_origins = TRUE,
                                       top_n_origins = 10,
                                       group_by_origin = FALSE,
                                       group_by_year = TRUE,
                                       group_by_commodities = FALSE,
                                       truncate_commodities = FALSE,
                                       top_n_commodities = 10,
                                       group_by_taxon = TRUE,
                                       truncate_taxon = TRUE,
                                       top_n_taxon = 10)

save_summary(data = diagnostic_results_summary2,
             base_filename = paste0(base_filename,'_plot2')
             )




save_plot(data = diagnostic_results_summary,
                      base_filename = paste0(base_filename,'_plot2'),
                      plot_x = "INSPECTION_YEAR",
                      xlab = "Year",
                      plot_y = "total_pests",
                      ylab = "Total Pests",
                      plot_fill = "PEST_TAXONOMIC_NAME",
                      fill_lab = "Pest taxonomic name",
                      plot_title = "Pest Interceptions by Year and Origin",
                      plot_subtitle = "Summary of pest interceptions",
                      width = 10,
                      height = 7,
                      dpi = 300)
```

```{r plt 3}

diagnostic_results_summary3 = summarize_diagnostic_results(df_diagnostic_results,
                                       arrange_by = "origin", #or "year"; anything else will be by total pests
                                       truncate_origins = TRUE,
                                       top_n_origins = 5,
                                       group_by_origin = TRUE,
                                       group_by_year = FALSE,
                                       group_by_commodities = FALSE,
                                       truncate_commodities = FALSE,
                                       top_n_commodities = 10,
                                       group_by_taxon = TRUE,
                                       truncate_taxon = TRUE,
                                       top_n_taxon = 15)

save_summary(data = diagnostic_results_summary3,
             base_filename = paste0(base_filename,'_summary_3')
             )




save_plot(data = diagnostic_results_summary3,
                      base_filename = paste0(base_filename,'_plot_3'),
                      plot_x = "ORIGIN",
                      xlab = "Origin",
                      plot_y = "total_pests",
                      ylab = "Total Pests",
                      plot_fill = "PEST_TAXONOMIC_NAME",
                      fill_lab = "Pest taxonomic name",
                      plot_title = "Pest Interceptions by origin and pest taxonomic name",
                      plot_subtitle = "Summary of pest interceptions",
                      width = 10,
                      height = 7,
                      dpi = 300)
```
