---
title: "copitarsia from anywhere"
author: "Dalon White"
date: "2025-08-21"
output: html_document

params:
  begin_date: "01-01-2020"
  end_date: "today" #"today" for all days, otherwise "mm-dd-yyyy"
  
  # Commodity taxonomy:
  taxonomic_name: NULL #NULL or ["any taxonomic level to match", "any taxonomic level to match", "etc."]
  common_name: NULL #NULL or ["corn", "sorghum", "etc."]
  
  # Country of origin:
  origin_country: NULL #NULL or ["country1", "country2", "etc."]
  
  # Pest taxonomy:
  KINGDOM: NULL #["Fungi"]
  PHYLUM: NULL
  CLASS: NULL
  ORDER: NULL
  FAMILY: NULL
  GENUS: ["Copitarsia"]
  SPECIES: ["decolora","corruda","incommoda","gibberosa"]
  
  # For the title of the file on save:
  pest_nickname: ["Copitarsia"]
  
---

# Add in flexibility to handle pest-specific names
# Change commmon_name and taxonomic_name to specific 'commodity' within them



```{r load libraries and grab params}
#get start and end dates
begin_date <- if(is.na(params$begin_date)) as.Date("1900-01-01") else as.Date(params$begin_date, "%d-%m-%Y")
end_date <- if(params$end_date == "today") Sys.Date() else as.Date(params$end_date, "%d-%m-%Y")

#get taxonomic names and countries
# taxonomic_name = params$taxonomic_name
# common_name = params$common_name
# origin_country = params$origin_country
# pest_taxonomic_name = params$pest_taxonomic_name
# pest_common_name = params$pest_common_name


#load packages
pacman::p_load(DBI, odbc)
pacman::p_load(tidyverse, lubridate, readr, stringr,stringdist)
pacman::p_load(DT, htmlwidgets)

```


# Gather data from ARM

## Connect to ARM
```{r connect to ARM}

connect_db <- function() {
  DBI::dbConnect(
    odbc::odbc(),
    .connection_string =
      "Driver=SQL Server;
                       Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                        Database=PPQ_AQI_ARMDMV2;
trusted_connection=yes")
}

db_conn <- connect_db()

```

## Grab diagnostic results table


```{r grab arm results}

mvw_diagnostic_results <- tbl(
  db_conn,
  sql("select * FROM [APHIS_Imports].[dbo].[mvw_Diagnostic_Results]")
)

df_diagnostic_results <- mvw_diagnostic_results
```
# Filter by date
```{r}
# Apply date filters only when needed
if (!is.null(begin_date)) {
  df_diagnostic_results <- df_diagnostic_results |> 
    filter(INSPECTION_DATE >= begin_date)
}

if (!is.null(end_date)) {
  df_diagnostic_results <- df_diagnostic_results |> 
    filter(INSPECTION_DATE <= end_date)
}
```
# Filter by country
```{r}
# Then conditionally apply the origin country filter if it exists
if (!is.null(params$origin_country) && length(params$origin_country) > 0) {
  df_diagnostic_results <- df_diagnostic_results |> 
    filter(ORIGIN %in% params$origin_country)
}
```
# Identify columns to collect
```{r grab the arm data}
df_diagnostic_results <- df_diagnostic_results |>   
  dplyr::select(
    INSPECTION_DATE,
    COMMODITY_DISPLAY_NAME,
    COMMODITY_TAXONOMIC_DISPLAY_NAME,
    INSPECTION_PATHWAY,
    ORIGIN,
    SUBCATEGORY,
    INSPECTION_LOCATION_STATE_CODE,
    DETERMINATION_TYPE,
    any_of(starts_with("PEST_TAXONOMIC_")),
    QUARANTINE_RECOMMENDATION,
    DIAGNOSTIC_DETERMINATION_ID,
    DIAGNOSTIC_REQUEST_ID,
    any_of(c(starts_with("number_dead", ignore.case = TRUE),
           starts_with("number_alive", ignore.case = TRUE),
           starts_with("pest_", ignore.case = TRUE))
    )
  )

```

## filter dates & names
```{r filter dates and names in ARM}

#Filter by commodity name or taxonomic name
arm_results <- df_diagnostic_results |> 
  collect()


```
# Filter by commodity
```{r}
# Only apply filtering if at least one parameter is not NULL
if (!is.null(params$common_name) || !is.null(params$taxonomic_name)) {
  # Create patterns only from non-NULL parameters
  if (!is.null(params$common_name)) {
    common_pattern <- paste(params$common_name, collapse = "|")
    has_common_pattern <- TRUE
  } else {
    has_common_pattern <- FALSE
  }
  
  if (!is.null(params$taxonomic_name)) {
    taxonomic_pattern <- paste(params$taxonomic_name, collapse = "|")
    has_taxonomic_pattern <- TRUE
  } else {
    has_taxonomic_pattern <- FALSE
  }
  
  # Build the filter dynamically based on available patterns
  if (has_common_pattern && has_taxonomic_pattern) {
    arm_results <- arm_results |> filter(
      str_detect(COMMODITY_DISPLAY_NAME, common_pattern) |
      str_detect(COMMODITY_TAXONOMIC_DISPLAY_NAME, taxonomic_pattern)
    )
  } else if (has_common_pattern) {
    arm_results <- arm_results |> filter(
      str_detect(COMMODITY_DISPLAY_NAME, common_pattern)
    )
  } else if (has_taxonomic_pattern) {
    arm_results <- arm_results |> filter(
      str_detect(COMMODITY_TAXONOMIC_DISPLAY_NAME, taxonomic_pattern)
    )
  }
}

```
# Filter by pest name
```{r}
filter_by_taxonomy <- function(filtered_data, params, threshold = 0.9) {
  # Extract names of non-NULL taxonomy parameters
  non_null_params <- names(params)[!sapply(params, is.null)]
  
  # If there are no non-NULL parameters, return the original data
  if (length(non_null_params) == 0) {
    return(filtered_data)
  }
  
  # For each non-NULL parameter, filter the data
  for (param in non_null_params) {
    # Create the column name with prefix
    column_name <- paste0("PEST_TAXONOMY_", param)
    
    # Check if the column exists in filtered_data
    if (column_name %in% colnames(filtered_data)) {
      # Get the values to filter by
      filter_values <- params[[param]]
      
      # Create empty result set to collect matches
      level_matches <- data.frame()
      
      # Process each value in the parameter vector
      for (param_value in filter_values) {
        # Find matches for this taxonomic value using fuzzy matching
        matches <- filtered_data %>%
          mutate(
            similarity = sapply(
              .data[[column_name]], 
              function(x) {
                if (is.na(x)) return(0)
                1 - stringdist(tolower(x), tolower(param_value), method = "jw") / 
                  max(nchar(x), nchar(param_value))
              }
            )
          ) %>%
          filter(similarity >= threshold)
        
        # Add to our matches collection
        level_matches <- bind_rows(level_matches, matches)
      }
      
      # Replace filtered_data with the matches from this level
      if (nrow(level_matches) > 0) {
        filtered_data <- level_matches %>% 
          distinct() %>%
          select(-similarity)
      } else {
        # No matches found for this level
        print(paste0("No matches found for level: ", param))
        filtered_data <- filtered_data[0, ] # Return empty dataset with same structure
        return(filtered_data)
      }
    } else {
      print(paste0("Column ", column_name, " not found in data"))
    }
  }
  
  return(filtered_data)
}

```

```{r filter_data}
# Example usage
arm_results <- filter_by_taxonomy(arm_results, params, threshold = 0.99)

# Show results
print(paste0("Found ", nrow(arm_results), " records matching taxonomic criteria"))
head(arm_results)
```

## Filter interceptions to best known information

```{r correct pest query diagnostic}
#call diagnostic determination data; written flexibly because QUARANTINE_STATUS_MAINLAND just chagned to QUARANTINE_STATUS_CONUS and it differs by table
get.determinations <- function(connection) {
  # First, get all columns to avoid SQL errors if column names change
  result <- tbl(connection, sql("SELECT
                       *
                       FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_DIAGNOSTIC_DETERMINATION]")) |>
    collect()

  # Get the actual column names from the retrieved data
  actual_cols <- colnames(result)

  # Define mandatory columns we need
  base_cols <- c(
    "ID",
    "DIAGNOSTIC_REQUEST_ID",
    "PEST_TAXON_ID",
    "PEST_TAXON_SIMPLE_NAME",
    "PEST_TAXONOMIC_NAME",
    "ID_AUTHORITY",
    "DETERMINATION_TYPE_ID",
    "DETERMINED_BY_GROUP_ID",
    "DETERMINATION_DATETIME",
    "QUARANTINE_STATUS_HAWAII",
    "QUARANTINE_STATUS_PUERTO_RICO"
  )

  # Check which quarantine status columns exist and add them
  quarantine_mainland_col <- NULL
  if ("QUARANTINE_STATUS_MAINLAND" %in% actual_cols) {
    quarantine_mainland_col <- "QUARANTINE_STATUS_MAINLAND"
  } else if ("QUARANTINE_STATUS_CONUS" %in% actual_cols) {
    quarantine_mainland_col <- "QUARANTINE_STATUS_CONUS"
  }

  # Create the final column selection list
  select_cols <- c(base_cols, quarantine_mainland_col)
  select_cols <- select_cols[!is.null(select_cols)]

  # Select the needed columns
  result <- result |>
    dplyr::select(all_of(select_cols)) |>
    rename("DIAGNOSTIC_DETERMINATION_ID" = "ID")

  # If needed, standardize column names for downstream code
  if (!is.null(quarantine_mainland_col) && quarantine_mainland_col == "QUARANTINE_STATUS_MAINLAND") {
    result <- result |> rename("QUARANTINE_STATUS_CONUS" = "QUARANTINE_STATUS_MAINLAND")
  }

  return(result)
}

system.time(
  diag.determ.records <- get.determinations(connection=db_conn)
  )

#pull BRG_DIATNOSTIC_DETERMINATION_NOT_POSSIBLE_REASON table 
## Just done to attach DIAGNOSTIC_NOT_POSSIBLE_FLAG in the event that a later molecular analysis was done but failed at a better determination (per Andy Carmichael's instructions, 8/12/24 on Teams)
get.diag.determ.not.possible = function(connection,
                                        determ.ids){
  tbl(connection, sql("SELECT
                      *
                      FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_DIAGNOSTIC_DETERMINATION_NOT_POSSIBLE_REASON]")) |> 
    # dplyr::filter(DIAGNOSTIC_DETERMINATION_ID %in% determ.ids) |> 
    collect() |> 
    dplyr::select(ID,
                  DIAGNOSTIC_DETERMINATION_ID,
                  DIAGNOSTIC_DETERMINATION_NOT_POSSIBLE_REASON)
}

determ.not.possible.dat = get.diag.determ.not.possible(
  connection=db_conn) |>  
  filter(DIAGNOSTIC_DETERMINATION_ID %in% (
    diag.determ.records |> pull(DIAGNOSTIC_DETERMINATION_ID)
    )
    )


#merge these
diag.determ.records = diag.determ.records |>
  left_join(determ.not.possible.dat,
            by=c('DIAGNOSTIC_DETERMINATION_ID'))


#Filter to records that match the data request
diag.determ.records <- diag.determ.records|> 
  filter(DIAGNOSTIC_DETERMINATION_ID %in% (arm_results |> pull(DIAGNOSTIC_DETERMINATION_ID)))

arm_results = arm_results |> left_join(diag.determ.records,by=intersect(colnames(arm_results),colnames(diag.determ.records))) 

parse_final_determination = function(data) {
  data |> 
    group_by(DIAGNOSTIC_REQUEST_ID) |> 
    #Andy Carmichael says if that the most recent "DETERMINATION_CREATED_DATETIME" should be the most accurate, with very rare exceptions.  (The only time I can think of is if there was a molecular ID request, and it failed.)   In that case there DETERM_NOT_POSS_FLAG should be flagged.  
    #Filter for rows that are not a failed diagnostic determination
    dplyr::filter(is.na(DIAGNOSTIC_DETERMINATION_NOT_POSSIBLE_REASON)) |> 
    #Filter the determination type - final has the highest TYPE_ID number
    dplyr::filter(DETERMINATION_TYPE_ID==max(DETERMINATION_TYPE_ID)) |> 
    #Filter for who ID'd it - the group_ID is a hierarchy of expertise, so take the max ID
    dplyr::filter(DETERMINED_BY_GROUP_ID == max(DETERMINED_BY_GROUP_ID)) |> 
    #The most recent is in practice the most correct - sometimes a final ID will be corrected at a later date, or someone will review for training or spot checking
    dplyr::filter(DETERMINATION_DATETIME == max(DETERMINATION_DATETIME))
}

df_diagnostic_results <- parse_final_determination(arm_results)

print(paste0("Found ", nrow(df_diagnostic_results), " ultimate identifications matching criteria"))
print(head(df_diagnostic_results))
```

## Write ARM data
```{r write table}
output_dir <- here::here('.gitignored','output','data')
if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
setwd(output_dir)

dt <- datatable(df_diagnostic_results,
                extensions = 'Buttons',
                options = list(
                  dom = 'Bfrtip',
                  buttons = c('copy', 'csv', 'excel', 'pdf')
                ))

# Generate base filename without extension
base_filename <- paste0(
  if(!is.null(params$pest_nickname)) (paste(params$pest_nickname, collapse="_")) else (paste("all pests")),
  "_on_",
  if(!(is.null(params$common_name) | is.null(params$taxonomic_name)))(paste(
    if(!is.null(params$common_name)) (paste(params$common_name, collapse="_")),
    if(!is.null(params$taxonomic_name)) (paste(params$taxonomic_name, collapse="_")))) else ("anything"),
  "_from_",
  if(!is.null(params$origin_country)) (paste(params$origin_country, collapse="_")) else ("anywhere"),
  "_from_period_",
  if(!is.null(params$begin_date)) (params$begin_date) else ("anytime"),
  "_to_",
  if(!is.null(params$end_date)) (params$end_date) else ("anytime"))

# Try to save as HTML widget, fall back to CSV if error occurs
tryCatch({
  saveWidget(dt, 
             file = paste0(base_filename, ".html"),
             selfcontained = TRUE)
  message("Successfully saved interactive HTML widget")
}, error = function(e) {
  message("Error saving as HTML widget: ", e$message)
  message("File may be too large. Saving as CSV instead.")
  write.csv(df_diagnostic_results, 
            file = paste0(base_filename, ".csv"), 
            row.names = FALSE)
  message("Data saved as: ", paste0(base_filename, ".csv"))
})

```
## summarize by origin country
```{r}
test = df_diagnostic_results |> mutate(INSPECTION_YEAR = lubridate::year(INSPECTION_DATE)) |> group_by(INSPECTION_YEAR, ORIGIN, COMMODITY_DISPLAY_NAME, COMMODITY_TAXONOMIC_DISPLAY_NAME) |> summarise(total_pests = n() )
view(test)
output_dir <- here::here('.gitignored','output','data')
if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
setwd(output_dir)
write.csv(test,paste(Sys.Date(),'copitarsia summary.csv'))

library(ggplot2)
library(dplyr)
library(viridis)

df <- test
# Identify top 6 countries by total pest count
top_origins <- df %>%
  group_by(ORIGIN) %>%
  summarise(total_country_pests = sum(total_pests, na.rm = TRUE)) %>%
  arrange(desc(total_country_pests)) %>%
  slice_head(n = 6) %>%
  pull(ORIGIN)

# Create a new dataframe with top 6 countries and "Other"
pest_summary <- df %>%
  mutate(ORIGIN_grouped = ifelse(ORIGIN %in% top_origins, ORIGIN, "Other")) %>%
  group_by(INSPECTION_YEAR, ORIGIN_grouped) %>%
  summarise(total_pests_sum = sum(total_pests, na.rm = TRUE)) %>%
  ungroup()

# Re-order factor levels based on total pest counts
origin_order <- pest_summary %>%
  group_by(ORIGIN_grouped) %>%
  summarise(overall_total = sum(total_pests_sum)) %>%
  arrange(desc(overall_total)) %>%
  pull(ORIGIN_grouped)

# Set factor levels to maintain order in the plot
pest_summary$ORIGIN_grouped <- factor(pest_summary$ORIGIN_grouped, 
                                     levels = origin_order)

# Create the plot
p = ggplot(pest_summary, aes(x = factor(INSPECTION_YEAR), y = total_pests_sum, fill = ORIGIN_grouped)) +
  geom_col(position = "dodge") +
  scale_fill_viridis_d(option = "inferno") +
  labs(
    title = "Total Pests by Origin and Year",
    subtitle = "Top 6 Countries and Other",
    x = "Inspection Year",
    y = "Total Number of Pests",
    fill = "Origin Country"
  ) +
  theme_classic() +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

output_dir <- here::here('.gitignored','output','data')
if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
setwd(output_dir)
# Also save the plot as a high-quality PNG
ggsave(paste0(base_filename, "_plot.png"), 
       plot = p, 
       width = 10, 
       height = 7, 
       dpi = 300)
```